Parsing Postfix log files

### 2007-06-29 doc/logparser.tex
Wrote a large chunk of the parsing coverage section; line coverage is pretty much done, mail coverage is started but needs lots of work (requires changes and improvements to the parser). Documented new actions. Documented situations when smtpd stops logging. Added context free grammar section. Attempted to make Latex accept the URLs in the bibliography without complaining. Misc cleanup.

% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib,nottoc]{tocbibind}
% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.
\usepackage{hyperref}
\hypersetup{%
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
    breaklinks  = true,
}

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table.} will check
whether the IP address of the connected client matches the left hand side
of each line in the file and will return the right hand side of the first
matching line.  Other restrictions determine their result in different
ways, e.g. \texttt{reject\_rbl\_client some.rbl.com} checks whether the IP
address of the client is present in the listed RBL\@.

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser: rules identify the line and return some data,
which the algorithm (the YACC part) receives and uses when performing the
requested action.  Rules are solely concerned with identifying a line and
extracting data from it, whereas the algorithm's task is to follow the
journey each mail takes through Postfix, piecing the data together into a
coherent whole, saving it in a useful and consistent form, and performing
housekeeping duties.  Readers familiar with context-free grammars may find
Section~\ref{comparison against context-free grammars} helpful.

@@

\subsection{Comparison against context-free grammars}

@@

XXX DESCRIBE IT IN LEFT\@: RIGHT PRODUCTION TERMS, IF POSSIBLE\@.

@@

    \item [rule\_order] is an efficiency measure.  This counter is
        maintained for every rule and incremented each time the rule
        successfully matches.  At the start of each run the program sorts
        the rules in descending rule\_order, and at the end of the run
        updates every rule's rule\_order.  Assuming that the distribution
        of log lines is reasonably consistent between log files, rules
        matching more commonly occurring log lines will be tried before
        rules matching less commonly occurring log lines, lowering the
        program's execution time.  Rule efficiency concerns are discussed
        in section~\ref{rule ordering for efficiency}.

@@

@@

@@

    \item The message-id has a specific format: \newline
        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
        e.g. \texttt{20070321125732.D168138A1@smtp.cs.tcd.ie}

@@

@@

XXX THIS SECTION NEEDS TO BE WRITTEN

@@

@@

>>>>

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
\usepackage{url}
\DeclareUrlCommand\urlwrap{}
\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib,nottoc]{tocbibind}
% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{%
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
    breaklinks  = true,
}

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table.} will check
whether the IP address of the connected client matches the left hand side
of each line in the file and will return the right hand side of the first
matching line.  Other restrictions determine their result in different
ways, e.g.\ \texttt{reject\_rbl\_client some.rbl.com} checks whether the IP
address of the client is present in the listed RBL\@.

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser: rules identify the line and return some data,
which the algorithm (the YACC part) receives and uses when performing the
requested action.  Rules are solely concerned with identifying a line and
extracting data from it, whereas the algorithm's task is to follow the
journey each mail takes through Postfix, piecing the data together into a
coherent whole, saving it in a useful and consistent form, and performing
housekeeping duties.  Readers familiar with context-free grammars may find
section~\ref{comparison against context-free grammars} helpful.

@@

@@

In context-free grammar terms the parser rules could be described as:

$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$

@@

    \item [rule\_order] is an efficiency measure.  This counter is
        maintained for every rule and incremented each time the rule
        successfully matches.  At the start of each run the program sorts
        the rules in descending rule\_order, and at the end of the run
        updates every rule's rule\_order.  Assuming that the distribution
        of log lines is reasonably consistent between log files, rules
        matching more commonly occurring log lines will be tried before
        rules matching less commonly occurring log lines, lowering the
        program's execution time.  Rule ordering for efficiency is
        discussed in section~\ref{rule ordering for efficiency}.

@@

    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
        it kills all smtpds,\footnote{Possibly other programs are killed
        also, but the parser is only affected by and interested in smtpd.}
        requiring any active connections to be committed and deleted from
        the state tables.

    \item [SMTPD\_KILLED] Sometimes Postfix needs to forcefully kill an
        smtpd; the active connection for that smtpd must be committed and
        deleted from the state tables.

    \item [SMTPD\_DIED] Sometimes an smtpd dies or exits unsuccessfully;
        the active connection for that smtpd must be committed and deleted
        from the state tables.

@@

\label{additional complications}

@@

    \item The message-id has a specific format: \newline
        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
        e.g.\ \texttt{20070321125732.D168138A1@smtp.cs.tcd.ie}

@@

\subsubsection{Smtpd stops logging}

\label{smtpd stops logging}

Occasionally an smtpd will just stop logging, without any explicit reason.
After poring over logs for some time there are several reasons for this
infrequent occurrence:

    \item Postfix is stopped or its configuration is reloaded.  When this
        happens all smtpds exit, and all entries in the connections state
        table must be committed and deleted from the state table.  The
        queueid state table is untouched.

    \item Postfix sometimes forcibly kills an smtpd, so the active
        connection must be committed and deleted from the connections state
        table.

    \item Occasionally an smtpd will die or exit uncleanly, so the active
        connection must be committed and deleted from the connections state
        table.

The above appear to cover all situations where an smtpd suddenly stops
logging.

@@

The discussion of the parser's coverage of Postfix log files is separated
into two parts: log lines covered and mails covered.  The first is
important because the parser should handle all (relevant) log lines it's
given; the second is equally important because the parser must properly
deal with every mail if it is to be useful.  Improving the former is
less intrusive, as it just requires new rules to be written; improving the
latter is much more intrusive as it requires changes to the parser
algorithm, and it can also be much harder to notice a deficiency.

@@

Parsing a log line is a three stage process:

    \item Check if there are any rules for the program which produced the
        log line; if not then skip the line.

    \item Try each rule, ordered by priority and rule order (see
        sections~\ref{rule attributes} and~\ref{rule ordering for
        efficiency}), until a matching rule is found.

    \item Execute the appropriate action.

Full coverage of log lines requires the following:

    \item Each program of interest must have at least one rule or it's log
        lines will be silently skipped; in the extreme case of zero rules
        the parser would happily skip every log line.  There may be any
        number of log lines from other programs intermingled in the log
        file, and there are some Postfix programs which don't produce any
        log lines of interest (currently \textit{postfix/anvil\/} and
        \textit{postfix/scache\/}).

    \item There must be a rule to match each different line produced by
        each program; if a line is not successfully matched the parser will
        issue a warning.  Rules should be as specific and tightly bound as
        possible to ensure accurate parsing:\footnote{A rule which matches
        zero or more of any character will successfully parse every log
        line, but not in a meaningful way.} most log lines contain fixed
        strings and have a rigid pattern, so this is not a problem.

    \item The appropriate action to take --- this is discussed in
        section~\ref{mails-covered}.

Full coverage of log lines is both easy to achieve and hard to maintain.
It is easy to achieve full coverage for a limited set of log files (at the
time of writing the parser has 131 rules, fully parsing 93 contiguous log
files from Postfix 2.2 and 2.3), and new rules are easy to add.
Maintaining full coverage is hard because other servers have different
restrictions with custom messages, RBLs change their messages over time,
major releases of Postfix change warning messages (usually adding more
information), etc.\ so over time the log lines drift and change.

XXX GRAPH RULE HITS

@@

Coverage of mails is much more difficult to determine accurately.  The
parser can dump its state tables in a human readable form; examining these
tables with reference to the log files is the best way to detect mails
which were not handled properly (all the complications discussed in
section~\ref{additional complications} were detected in this way).  The
parser issues warnings when it detects any errors, some of which may alert
the user to a problem, e.g.\ when a queueid is reused before the previous
mail is fully dealt with, when a queueid/pid is not found, or when there
are problems tracking a child mail (see
section~\ref{tracking-re-injected-mail}).  There should be few or no
warnings (warnings about mail where some of the log lines are not in the
logfile can be ignored), and the state table should only contain entries
for mails which had yet to be delivered when the logfiles ended.

At the time of writing the parser is being tested with 93 log files.  The
state table contains entries for mails not yet delivered at the end of the
